Complex, informative, and previously ignored dependencies propogate throughout ecology and can be revealed using copulas

### 2019-05-10 Paper.Rmd

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird body masses and BMR", "Mammal body masses and BMR", 
                      "Green spruce and other aphid species abundances", "Leaf-curling plum and other aphid first flight dates","Ceratium furca and other plankton taxa abundances", "Plant diversity and aboveground biomass","Methane flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait",
                          "Population","Phenological","Population","Community level","Ecosystem functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "14 locations with at least 45-yr. timeseries for each 22 taxa",
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "13 locations with at least 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab (2009)","McNab (2008)",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Cedar Creek Ecosystem Science Reserve, biodiversity experiment e120",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

knitr::kable(tab_data_info, 
             format="latex", align="l",
             caption = "Summary table for the data we used. Bold entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(c(4,5,6,8), bold=T)
```

@@

We present green spruce aphid abundance results (Table \ref{tab_data_info})
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-N
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than N-copula weights (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of OH. Goodness of fit tests 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data; the collection of 
copula families we used was apparently sufficiently broad that all pairs of 
locations could be characterized by at least one of them.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
OH. These data have greater tail dependence 
(lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower- minus upper-tail dependence statistics 
(Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of OH.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--***DAN: You could maybe move the next para to sup mat, though that would complicate the
presentation of skewness results below. Only do it if you are desperate to cut.-->
<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes OH results from
multivariate data. Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown); conclusions supported that 
non-N copula structure and tail dependence, and asymmetry of tail dependence, are 
common.

>>>>

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird body masses and BMR", "Mammal body masses and BMR", 
                      "Green spruce and other aphid species abundances", "Leaf-curling plum and other aphid first flight dates","\\textit{Ceratium furca} and other plankton taxa abundances", "Plant diversity and aboveground biomass","Methane-flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait",
                          "Population","Phenological","Population","Community level","Ecosystem functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "14 locations with at least 45-yr. timeseries for each 22 taxa",
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "13 locations with at least 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab (2009)","McNab (2008)",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Cedar Creek Ecosystem Science Reserve, biodiversity experiment e120",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

knitr::kable(tab_data_info, 
             #format="latex", align="l",
             format="latex", align="l",escape=F,
             caption = "Summary table for the data we used. Bold entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(c(4,5,6,8), bold=T)
```

@@

We present green spruce aphid abundance results (Table \ref{tab_data_info})
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-N
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than N-copula weights (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of OH. Goodness of fit tests 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data; the collection of 
copula families we used was apparently sufficiently broad that all pairs of 
locations could be characterized by at least one of them.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
OH. These data have greater tail dependence 
(lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower- minus upper-tail dependence statistics 
(Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few location pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of OH.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--***DAN: You could maybe move the next para to sup mat, though that would complicate the
presentation of skewness results below. Only do it if you are desperate to cut.-->
<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes OH results from
multivariate data. Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most location pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane-flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown); conclusions supported that 
non-N copula structure and tail dependence, and asymmetry of tail dependence, are 
common.
